---
title: "Empirical Application in Gunsilius (2023)"
author: "David Van Dijcke"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pre-Testing in a DiD Setup using the did Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: ../inst/REFERENCES.bib
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    comment = "#",
    echo=TRUE,
  error = FALSE,
  tidy = FALSE,
  cache = FALSE,
  collapse = TRUE,
  out.width = '100%',
  dpi = 144
)
```

```{r, include=FALSE}

# fldr <- here::here("R/")
# sapply(paste0(fldr,list.files(fldr)), source)# library(CVXR)
library(pracma)
library(quadprog)
library(parallel)
library(stats)
library(data.table)
library(CVXR)
library(DiSCo)
# fldr <- here::here("R/")
# sapply(paste0(fldr,list.files(fldr)), source)
```

# Introduction

This vignette demonstrates how to use the DiSCo package by way of the
empirical application in @gunsilius2023distributional, which is based on
@dube2019minimum. We illustrate the use of the two main functions: 1)
`DiSCo`, which estimates the raw distributional counterfactuals,
computes confidence intervals, and optionally performs a permutation
test; and 2) `DiSCoTEA`, the "Treatment Effect Aggregator", which takes
in the distributional counterfactuals and computes aggregate treatment
effects using a user-specified aggregation statistic.

# Distributional Synthetic Controls

We briefly review the main idea behind Distributional Synthetic
Controls. Denote $Y_{jt,N}$ the outcome of group $j$ in time period $t$
in the absence of an intervention. Also denote $Y_{jt,I}$ the outcome in
the presence of an intervention at time $t > T_0$. Denote the quantile
function, $$
F^{-1}(q):=\inf _{y \in \mathbb{R}}\{F(y) \geq q\}, \quad q \in(0,1),
$$ where $F(y)$ is the corresponding cumulative distribution function.
One unit $j=1$ has received treatment, while the other units
$j=2,\ldots,J$ have not. Then the goal is to estimate the counterfactual
quantile function $F_{Y_{1 t}, N}^{-1}$ of the treated unit had it not
received treatment by an optimally weighted average of the control
units' quantile functions, $$
F_{Y_{1 t}, N}^{-1}(q)=\sum_{j=2}^{J+1} \lambda_j^* F_{Y_{j t}}^{-1}(q) \quad \text { for all } q \in(0,1)
$$\
In practice, we do this by solving the following optimization problem:
$$
\vec{\lambda}_t^*=\underset{\vec{\lambda} \in \Delta^J}{\operatorname{argmin}} \int_0^1\left|\sum_{j=2}^{J+1} \lambda_j F_{Y_{j t}}^{-1}(q)-F_{Y_{1 t}}^{-1}(q)\right|^2 d q
$$ which gives optimal weights for each period. This problem can be
solved by a simple weighted regression, which is implemented in the
`DiSCo_weights_reg` function. To obtain the overall optimal weights
$\vec{\lambda}^*$, we take the average of the optimal weights across all
periods.

# Illustration

The data from @dube2019minimum is available in the package. We load it
here:

```{r}
data("dube")
head(dube)
```

To learn more about the data, just type `?dube` in the console. We have
already renamed the outcome, id, and time variables to `y_col`,
`id_col`, and `time_col`, respectively, which is required before passing
the dataframe to the DiSCo command. We also need to set the two
following parameters:

```{r}
id_col.target <- 2
t0 <- 2003
```

which indicate the id of the treated unit and the time period of the
intervention, respectively. We can now run the DiSCo command:

```{r}
# if the below gives issues it's the knit cache...
disco <- DiSCo(dube, id_col.target, t0, M = 1000, G = 1000, num.cores = 5, permutation = FALSE,
                 CI = TRUE, boots = 1000, cl = 0.95, CI_placebo=TRUE, graph = FALSE, qmethod=NULL)
```

where we have chosen grids of 1000 grid points and opted for parallel
computation with 5 cores to speed up the permutation test and confidence
intervals, which we calculate using 500 resamples.

We can use the DiSCo Treatment Effect Aggregator (`DiSCoTEA`) function
to aggregate the resulting counterfactual quantile functions into
various treatment effect measures. For example, we can calculate the
average treatment effect (ATE) as follows:

```{r}
DiSCoTEA(disco,  agg="ATT", graph=TRUE, time=TRUE)

```

We can also look at the CDF of the treatment effects,

```{r}
DiSCoTEA(disco,  agg="cdfTreat", graph=TRUE, time=TRUE, n_per_window=NULL)

```

Or the quantile function:

```{r}
DiSCoTEA(disco,  agg="quantileTreat", graph=TRUE, time=TRUE, n_per_window=NULL)
```

To inspect what is happening at the higher quantiles, where we appear to
see an increase in family income as a result of the minimum wage policy,
we can separately plot the observed and counterfactual functions,

```{r}
DiSCoTEA(disco,  agg="quantile", graph=TRUE, time=TRUE, n_per_window=NULL)

```

# Permutation test

Similar to the canonical permutation test in @abadie2010synthetic, we
can carry out a distributional permutation test. The idea is entirely
analogous: we iteratively reassign the treatment to units in the set of
controls and estimate "placebo" quantile functions. Then we calculate
the squared Wasserstein error of the entire distribution as, $$
d_{t t}^2:=\int_0^1\left|F_{Y_{u t, N}}^{-1}(q)-F_{Y_{u t}}^{-1}(q)\right|^2 d q
$$ for each time period. The `permutation` option in the `DiSCo`
function allows one to plot the time evolution of these terms for all
units by setting `graph = TRUE`, to visually inspect the abnormality of
the "true" treatment effects. We also calculate the p-value of the test
as, $$
p_t=\frac{r\left(d_{1 t}\right)}{J+1}
$$ where $r(d_{1t})$ is the rank of the true treatment effect in the
distribution of placebo effects.

```{r}
results <- DiSCo(dube, id_col.target, t0, M = 1000, G = 1000, num.cores = 5, permutation = TRUE, CI = FALSE, boots = 1000, cl = 0.95, CI_placebo=TRUE, graph = TRUE, qmethod=NULL) 

```
